Local real space operators for translating Stokes Q/U to E/B maps on the sphere

### 2017-09-19 article/sections/numerical_implementation.tex
reorganized the introduction sections ; completed section on numerical implementation qu2eb

At first glance, the above remarks seem to hold true only for the E mode spectra, since the errors in the B mode spectra seem to be significantly larger as seen in the middle row of \fig{fig:eb-spectra_rad_cutoff}. We understand that the higher error in the B mode spectra is primarily a combined effect of the low amplitude of the B mode spectrum as compared to the E mode spectrum and limited numerical accuracy of our simple convolution algorithm. We ran a number of tests to confirm these as reasons for the observed discrepant behaviour. We exchanged the E and B mode spectra, \revisit{(taking care to set $C_{\ell}^{TE}=0$)} and noted that the trends observed in the error plots are reversed, clearly indicating a dependence on the amplitude of the spectrum. We noted that the polar caps were particularly error prone, as seen in \fig{fig:eb-maps-compare}. We excised these error prone regions, specifically masking portions of the maps with $|b|>30^{\circ}$ to find that the errors on the B mode spectra revert to having very similar structure as that of the E mode spectra, as seen in the bottom row of \fig{fig:eb-spectra_rad_cutoff}. Finally we modified the 
%
\begin{figure}[!h] 
\centering
%\vspace{-5cm}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/bb-spectrum-radial-cutoff.pdf}}\\[-6ex]
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-bb-spectrum-radial-cutoff.pdf}}\\[-6ex]
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/eqmask/relative-percentage-err-ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/eqmask/relative-percentage-err-bb-spectrum-radial-cutoff.pdf}}
\caption{\textit{Top:} The figures on left and right depict the E mode spectra and the B mode spectra respectively. \textit{Middle:} These figures depict the relative percentage difference between the spectra derived from full sky, locally evaluated E \& B  maps and the reference spectra. \textit{Bottom:} These figures depict the same results as shown in the middle row, except that the regions with $|b|>30^\circ$ of the maps have been masked for this analysis. 
The colors indicate the spectra derived from imposing different radial cutoff as indicated by the legends. The gray bands indicate the cosmic variance scatter around the fiducial spectra. The dashed horizontal black lines mark the 2\% error.}
\label{fig:eb-spectra_rad_cutoff}
\end{figure}
%
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Separating Stokes Q \& U maps corresponding to E \& B modes of polarization}
%
\begin{figure}[!h] 
\centering
\subfigure[E-Q map Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-healpix.pdf}}
\subfigure[E-Q map local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-2beta.pdf}}
\subfigure[E-Q map difference]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-diff.pdf}}
\subfigure[E-U map Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-healpix.pdf}}
\subfigure[E-U map local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-2beta.pdf}}
\subfigure[E-U map difference]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-diff.pdf}}

\subfigure[B-Q map Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-healpix.pdf}}
\subfigure[B-Q map local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-2beta.pdf}}
\subfigure[B-Q map difference]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-diff.pdf}}
\subfigure[B-U map Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-healpix.pdf}}
\subfigure[B-U map local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-2beta.pdf}}
\subfigure[B-U map difference]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-diff.pdf}}
\caption{}
\label{fig:equ-bqu-maps-compare}
\end{figure}
%

%
\begin{figure}[!h] 
\centering
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/equ-spectra-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/bqu-spectra-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-equ-e-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-bqu-b-spectrum-radial-cutoff.pdf}}
\caption{}
\label{fig:equ-bqu-spectra_rad_cutoff}
\end{figure}
%
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Multipole filtering using spatial convolutions}
%--------------------------------------------------------
%--------------------------------------------------------

>>>>

At first glance, the above remarks seem to hold true only for the E mode spectra, since the errors in the B mode spectra seem to be significantly larger as seen in the middle row of \fig{fig:eb-spectra_rad_cutoff}. We understand that the higher error in the B mode spectra is primarily a combined effect of the low amplitude of the B mode spectrum as compared to the E mode spectrum and limited numerical accuracy of our simple convolution algorithm. We ran a number of tests to confirm these as reasons for the observed discrepant behaviour. 
%
\begin{figure}[!t] 
\centering
%\vspace{-5cm}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/bb-spectrum-radial-cutoff.pdf}}\\[-6ex]
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-bb-spectrum-radial-cutoff.pdf}}\\[-6ex]
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/eqmask/relative-percentage-err-ee-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/eqmask/relative-percentage-err-bb-spectrum-radial-cutoff.pdf}} \\[-2ex]
\caption{\textit{Top:} The figures on left and right depict the E mode spectra and the B mode spectra respectively. \textit{Middle:} These figures depict the relative percentage difference between the spectra derived from full sky, locally evaluated E \& B  maps and the reference spectra. \textit{Bottom:} These figures depict the same results as shown in the middle row, except that the regions with $|b|>30^\circ$ of the maps have been masked for this analysis. 
The colors indicate the spectra derived from imposing different radial cutoff as indicated by the legends. The gray bands indicate the cosmic variance scatter around the fiducial spectra. The dashed horizontal black lines mark the 2\% error.}
\label{fig:eb-spectra_rad_cutoff}
\end{figure}
%
We exchanged the E and B mode spectra, \revisit{(taking care to set $C_{\ell}^{TE}=0$)} and noted that the trends observed in the error plots are reversed, clearly indicating a dependence on the amplitude of the spectrum. We noted that the polar caps were particularly error prone, as seen in \fig{fig:eb-maps-compare}. We excised these error prone regions, specifically masking portions of the maps with $|b|>30^{\circ}$ to find that the errors on the B mode spectra revert to having very similar structure as that of the E mode spectra, as seen in the bottom row of \fig{fig:eb-spectra_rad_cutoff}. Finally we modify the use of standard Healpix routines to mimic the local convolution E \& B construction (see \app{sec:convolution_err} for details). The results from this exercise also concur with our understanding of the discrepant behaviour for the B-mode spectrum.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Separating Stokes Q \& U maps corresponding to E \& B modes of polarization}
%
\begin{figure}[!h] 
\centering
\subfigure[E Q-map: Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-healpix.pdf}}
\subfigure[E Q-map: Local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-2beta.pdf}}
\subfigure[E Q-map: Difference]{\includegraphics[width=0.31\columnwidth]{simulated/e-qmap-diff.pdf}}
\subfigure[E U-map: Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-healpix.pdf}}
\subfigure[E U-map: Local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-2beta.pdf}}
\subfigure[E U-map: Difference]{\includegraphics[width=0.31\columnwidth]{simulated/e-umap-diff.pdf}}

\subfigure[B Q-map: Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-healpix.pdf}}
\subfigure[B Q-map: Local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-2beta.pdf}}
\subfigure[B Q-map: Difference]{\includegraphics[width=0.31\columnwidth]{simulated/b-qmap-diff.pdf}}
\subfigure[B U-map: Healpix]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-healpix.pdf}}
\subfigure[B U-map: Local convolution]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-2beta.pdf}}
\subfigure[B U-map: Difference]{\includegraphics[width=0.31\columnwidth]{simulated/b-umap-diff.pdf}}
\caption{The top two rows depict the Stokes Q \& U maps  respectively corresponding to E mode of polarization while the bottom two rows depict  the Stokes Q \& U maps  respectively corresponding to the  B modes of polarization. \textit{Left:} Stokes Maps derived using Healpix. \textit{Middle}: Stokes maps derived using local convolution with $r_{\rm cutoff} = 2 \beta_o$. \textit{Right:} Difference between the maps derived using the two different methods.}
\label{fig:equ-bqu-maps-compare}
\end{figure}
%

%
\begin{figure}[!h] 
\centering
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/equ-spectra-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/bqu-spectra-radial-cutoff.pdf}}\\[-5ex]
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-equ-e-spectrum-radial-cutoff.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{simulated/relative-percentage-err-bqu-b-spectrum-radial-cutoff.pdf}}
\caption{}
\label{fig:equ-bqu-spectra_rad_cutoff}
\end{figure}
%
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Multipole filtering using spatial convolutions}
%--------------------------------------------------------
%--------------------------------------------------------

### 2017-09-19 article/sections/real_space_formalism.tex
reorganized the introduction sections ; completed section on numerical implementation qu2eb

\subsection{Matrix notation} \label{sec:mat_pol_intro}
In this section we cast the relation introduced in Sec.~\ref{sec:pol-primer} in matrix notation\footnote{While we work with the matrix and vector sizes given in terms of Healpix pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}. This representation will make transparent the derivation of the real space operators we discuss in the following sections. We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.\\
We begin by introducing the matrices encoding the spin spherical harmonic basis vectors,
%
\beq
_sB= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,,
\eeq
%
where $s$ denotes the spin of the basis functions. For this work we will only be working with cases $s \in [0,2]$. In this notation, each column can be mapped to a specific harmonic basis function marked by the pair of indices:$(\ell,m)$ and each row maps to a specific position on the sphere. Note that this matrix is in general not a square matrix. Generally the number of columns is determined by the scheme used to discretely represent the sphere and the number of rows is set by the number of basis functions of interest (often determined the band limit).

>>>>

%--------------------------------------------------------
%--------------------------------------------------------
\section{E/B description of CMB polarization} \label{sec:pol-intro}
%--------------------------------------------------------
\subsection{Polarization primer}\label{sec:pol-primer}
The CMB polarization is measured in terms of Stokes Q and U parameters. These measurements can be combined to form the complex spin 2 polarization field as follows,
%
\beqry \label{eq:spin-pol}
_{\pm 2}\bar{X}(\hat{n}) &=& Q(\hat{n}) \pm i U (\hat{n}) \nonumber \\ &=& \sum_{\ell m}  {_{\pm 2}} \tilde{X}_{\ell m}  {_{\pm 2}}Y_{\ell m} (\hat{n}) \,.
\eeqry
%
Since these measured quantities depend on the local coordinate system making it is cumbersome to work with them. To overcome this, one describes the CMB polarization field in terms of a scalar field denoted by $E(\hat{n}) $ and a pseudo scalar field $B(\hat{n}) $ \cite{Kamionkowski1997}. These scalar fields are related to the spin-2 polarization field $_{\pm 2}X(\hat{n})$ through the following relation,
%
\beq \label{eq:ebdef}
\mathcal{E}(\hat{n}) = -\frac{1}{2} \big[ \bar{\eth}^2 _{+ 2}X(\hat{n})  +  \eth^2 _{- 2}X(\hat{n}) \big] ~\,;~\mathcal{B}(\hat{n}) = -\frac{1}{2i} \big[ \bar{\eth}^2 _{+ 2}X(\hat{n})  -  \eth^2 _{- 2}X(\hat{n}) \big] \,,
\eeq
%
where $\eth$ and $\bar{\eth}$ denote the spin raising and lowering operators respectively. These $E$ and $B$ fields are spin-0 fields similar to the temperature anisotropies and hence their value are independent of the coordinate system definitions. The spin raising and lowering operators have the following properties \cite{goldberg67},
%
\beqrys \label{eq:spinopylm} 
\eth _s Y_{lm}(\hat{n}) &=& \sqrt{(\ell-s)(\ell+s+1)} _{s+1} Y_{lm}(\hat{n}) \,, \\
\bar{\eth} _s Y_{lm}(\hat{n}) &=& -\sqrt{(\ell+s)(\ell-s+1)} _{s-1} Y_{lm}(\hat{n}) \,, 
\eeqrys
%
where $_s Y_{lm}(\hat{n}) $ denote the spin-s spherical harmonics.

Using \eq{eq:ebdef} and the properties of the spin raising and lowering operators given in \eq{eq:spinopylm} it can be shown that the scalar fields $\mathcal{E}/\mathcal{B}$ are defined via the following set of equations,
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) ~\,;~ \mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients of  $\mathcal{E}/\mathcal{B}$ fields are related to the spin harmonic coefficients of the polarization field through the following equations,
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}X_{\ell m} + _{-2}X_{\ell m} \Big] ~\,;~a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}X_{\ell m} - _{-2}X_{\ell m} \Big] 
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields as defined by the following expressions, 
%
\beq \label{eq:pseudo}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n}) ~\,;~ B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
Note that the two set of fields $\mathcal{E}/\mathcal{B}$ and $E/B$ differ, since their spherical harmonic coefficients of expansion differ by the factor of $\sqrt{\frac{(\ell+2)!}{(\ell-2)!}}$.

%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Matrix notation} \label{sec:mat_pol_intro}
In this section we cast the relation introduced in Sec.~\ref{sec:pol-primer} in matrix notation\footnote{While we work with the matrix and vector sizes given in terms of Healpix pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}. This representation will make transparent the derivation of the real space operators we discuss in the following sections. We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.\\
We begin by introducing the matrices encoding the spin spherical harmonic basis vectors,
%
\beq
_sB= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,,
\eeq
%
where $s$ denotes the spin of the basis functions. For this work we will only be working with cases $s \in [0,2]$. In this notation, each column can be mapped to a specific harmonic basis function marked by the pair of indices:$(\ell,m)$ and each row maps to a specific position on the sphere. Note that this matrix is in general not a square matrix. Generally the number of columns is determined by the scheme used to discretely represent the sphere and the number of rows is set by the number of basis functions of interest (often determined the band limit).
